---
title: |
 | Statistical Inference
 | Simulation of the Exponential Distribution
author: "Gabor Simon"
output: 
  pdf_document:
          toc: true
---


## Overview

We are investigating the exponential distribution and the way the Central Limit Theorem
applies to it.

For this, we use an exponential distribution with a rate parameter of $\lambda$ = 0.2,
we take 40 samples in each simulation step, and we perform 1000 simulations.

A quick intro for the exponential distribution can be found on [Wikipedia](https://en.wikipedia.org/wiki/Exponential_distribution),
but the most important for us now is that

* its mean is $\mu = \frac{1}{\lambda}$ (in our case 1 / 0.2 = 5) and 
* its standard deviation is also $\sigma = \frac{1}{\lambda}$ (in our case also 5), so
* its variance is the square of that $\sigma^{2} = \frac{1}{\lambda^{2}}$, (in our case $5^{2} = 25$)


So, our R code begins with setting the parameters:
```{r set_params, message = FALSE}
library(ggplot2)
# We want the same data on each run
set.seed(1507446545)

# For all simulations, the parameter of the exponential is this value
lambda <- 0.2

# We take sets of this many samples
n_samples <- 40

# Number of simulations
n_simulations <- 1000
```


## Simulations
To simulate a phenomenon that behaves according to an exponential distribution, we 
generate a lot of random numbers by using the `rexp()` function.

We arrange these values in a matrix, where each row represents a simulation,
so this matrix has as many columns as many samples we take in one simulation.

```{r generate_data, message = FALSE}
# Each row represents a simulation of `n_samples` samples
samples <- matrix(rexp(n_samples * n_simulations, rate = lambda),
		  nrow = n_simulations)
dim(samples)

# Now we calculate the means of the samples
sample_means <- apply(samples, 1, mean)
head(sample_means)

# And we can plot their histogram
g_hist <- geom_histogram(aes(sample_means),
			 bins = 30,
			 colour = "black",
			 fill = "white")
g_pop_mean <- geom_vline(xintercept = 1 / lambda,
			 colour = "blue")

ggplot() +
	g_hist +
	g_pop_mean +
	labs(title = "Sample mean", x = "mean value", y = "frequency")
```

The histogram above shows the frequencies of the *sample* means, at first glance it looks like kind of a
normal distribution that is centered around the supposed population mean 5 (blue line).


## Sample Mean vs. Theoretical Mean
To get more accurate parameters about the distribution of the sample means, let's check their mean first:
```{r mean_of_sample_means, message = FALSE}
mx = mean(sample_means)
mx
```

So the means are indeed centered around 5, which they should be according to the Central Limit Theorem:

*The means of the sample means estimate the population mean*

In our case, the value around 5 above estimates $\mu = 5$.


## Sample Variance vs. Theoretical Variance
To continue checking the distribution of the sample means, let's check their variance as well:

```{r var_of_sample_means, message = FALSE}
vx = var(sample_means)
vx
```

The means have a variance about 0.63 (so their standard deviation is about 0.8).

According to the Central Limit Theorem:

*The variance of the sample means estimate the population variance divided by the sample size.*

In our case, the value around 0.63 above estimates $\frac{\sigma^{2}}{n_{samples}} = \frac{25}{40} = 0.625$, which is quite close.

## Distribution of the Means

If we approximate the *probability* of getting a mean less than a limit by the
*ratio* of such means among our simulations, then we can define the distribution
function of the sample means:

```{r dist_fn_of_sample_means, message = FALSE}
psm <- Vectorize(function(x) sum(sample_means < x) / n_simulations)
```
 
We suspect that this distribution approximates $Z(mean=1/\lambda, sd=\frac{1/\lambda}{\sqrt{n_{samples}}})$:

```{r compare_dist_fns, message = FALSE}

g_psm <- stat_function(fun = psm,
		       aes(colour = "Actual"))

g_norm <- stat_function(fun = function(x) pnorm(x,
						mean = 1/lambda,
						sd = 1/lambda / sqrt(n_samples)),
			aes(colour = "Normal"))

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
	xlim(2, 8) + ylim(0, 1) +
	g_psm + g_norm +
	labs(title = "Distribution functions", x = "x", y = "P(mean < x)") +
	scale_colour_manual("Dist. fn.",
			    values = c("Actual" = "red", "Normal" = "blue"))
```

It seems to fit but how closely? As the distribution of the means is somewhat
mound-shaped, we may calculate a p-value by using a classic 'treated vs. placebo' t-test,
where our mean values are the 'treated' set, and a normal distribution is the the 'placebo'.

Our null hypothesis is that the means follow the Central Limit Theorem, that is,
they are $Z(mean=1/\lambda, sd=(1/\lambda)/\sqrt{n_{samples}})$.

```{r p_value, message = TRUE}
# The parameters of the 'treated' set:
mx <- mx
sx <- sqrt(vx)
nx <- n_simulations
c(mx, sx, nx)

# The parameters of the 'placebo' set:
my <- 1/lambda
sy <- 1/lambda / sqrt(n_samples)
ny <- n_simulations # same size
c(my, sy, ny)

# The pooled variance estimator:
sp <- sqrt( ((nx-1)*sx^2 + (ny-1)*sy^2) / (nx + ny - 2) )
sp

# The test statistic:
ts <- (mx - my) / (sp * sqrt(1/nx + 1/ny))
ts

# The p-value:
# (Checking for both tails around the confidence interval, so calculate the
# left tail and double it, because the right tail covers the same.)
p <- 2 * pt(-abs(ts), nx + ny - 2)
p
```
This p-value is definitely high enough, it means that according to the null hypothesis it is
quite likely to get this result, so we **fail to reject** the null hypothesis, so we may
assume that the means indeed follow the Central Limit Theorem.

## Distribution of the Exponentials

It is important that it is the **sample means** that converge to a normal distribution,
although the original data follows the exponential. To demonstrate this, here is the
histogram of the same original data:

```{r orig_histogram, message = TRUE}
g_hist <- geom_histogram(aes(as.vector(samples)),
			 bins = 30,
			 colour = "black",
			 fill = "white")
g_pop_mean <- geom_vline(xintercept = 1 / lambda,
			 colour = "blue")

ggplot() +
	g_hist +
	g_pop_mean +
	labs(title = "Original (exponentially distributed) data", x = "mean value", y = "frequency")
```




